---
title: "ENV S 193DS Final"
author: "Niki van der Poel"
format: pdf
date: June 9 2026
---

[Github Repo] (https://github.com/nvanderpoel-cmd/ENVS-193DS_spring-2026_final)

# Set up
```{r}
#| message: false
#| label: Set up

# reading in packages
library(here)
library(tidyverse)
library(janitor)
library(dplyr)
library(MuMIn)
library(DHARMa)
library(ggeffects)
library(ggplot2)

# reading in data
nests <- read.csv(here("data/nest_data_final.csv"))

```

# Problem 1. Research Writing

## a. Transparent statistical methods

## b. Suggestions for rewriting

## c. Further analysis

# Problem 2. Data analysis

## a. Clean and wrangle
```{r}
#| label: clean and wrangle

# new object nests_clean
nests_clean <- nests |> 
  select(height_cm, case_control, ant_species) |> 
  filter(ant_species %in% c(
                     "AB",
                   "RRB",
                   "BBR")) |> 
  mutate(ant_species = as_factor(ant_species),
         ant_species = fct_relevel(ant_species,
                                    "AB",
                                    "RRB",
                                    "BBR")) |> 
  mutate(species_name = case_match(
    ant_species,
    "AB" ~ "Crematogaster sjostedti",
    "RRB" ~ "Crematogaster mimosae",
    "BBR" ~ "Crematogaster nigriceps"
  ))

slice_sample(nests_clean, n = 10)
```


## b. Response variable

For the response variable, case_control, the 1's signify there is a nest, of any bird species, present, while a 0 signifies there is no nest present. 

## c. Purpose of study

Tree height: Tree height is a continuous variable. As tree height increases, the probability of a nest occurrence would increase because of increased distance between nests and browsing mammals.

Acacia ant species: Acacia ant species is a categorical variable. 

## d. Table of models
| Model number | Tree Height | Ant Species | Interaction | 
|--------------|:-----------:|:-----------:|:-----------:| 
| 1            |      X      |      X      |     X       | 
| 2            |      X      |      X      |             |

## e. Run the models

```{r}
#| message: false
#| label: running models

# model 1: Model with interaction term

model1 <- glm(case_control ~ height_cm * ant_species,
              data = nests_clean,
              family = "binomial"
              )

# model 2: Model without interaction term

model2 <- glm(case_control ~ height_cm + ant_species,
              data = nests_clean,
              family = "binomial"
              )

```


## f. Select the best model

```{r}
#| label: model-selection

AICc(
model1,
model2
) |> 
arrange(AICc)
```

The best model as determined by Akaike's Information Criterion (AIC) was model 1, with interaction.

## g. Check the diagnostics

```{r}
#| label: check diagnostics

plot(
  simulateResiduals(model1)
)
```

## h. Visualize the model predictions

```{r}
#| label: model predictions

ggplot(nests_clean, aes(x = height_cm, y = case_control, color = species_name)) +
  geom_point(alpha = 0.3) +
  stat_smooth(method = "glm",
              method.args = list(family = "binomial"),
              se = TRUE) +
  facet_wrap(~species_name) +
  labs(
    x = "Tree height (cm)",
    y = "Probability of nest occurrence",
    color = "Ant species"
  ) +
  theme(legend.position = "none",
        panel.background = element_blank
        )
```


## i. Write a caption for your figure

## j. Calculate model predictions

## k. Interpret your results

# Problem 3. Affective and exploratory visualizations

## a. Comparing visualizations

## b. Designing an analysis

## c. Check any assumptions and run your analysis

## d. Create a visualization

## e. Write a caption

## f. Write about your results

## g. Sharing your affective visualization

















