---
title: "ENV S 193DS Final"
author: "Niki van der Poel"
format: pdf
date: June 9 2026
---

[Github Repo] (https://github.com/nvanderpoel-cmd/ENVS-193DS_spring-2026_final)

# Set up
```{r}
#| message: false
#| label: Set up

# reading in packages
library(here)
library(tidyverse)
library(janitor)
library(dplyr)
library(MuMIn)
library(DHARMa)
library(ggeffects)
library(ggplot2)

# reading in data
nests <- read.csv(here("data/nest_data_final.csv"))

```

# Problem 1. Research Writing

## a. Transparent statistical methods

## b. Suggestions for rewriting

## c. Further analysis

# Problem 2. Data analysis

## a. Clean and wrangle
```{r}
#| label: clean and wrangle

# new object nests_clean
nests_clean <- nests |> 
  select(height_cm, case_control, ant_species) |>  
  filter(ant_species %in% c(
                     "AB",
                   "RRB",
                   "BBR")) |> 
  mutate(ant_species = as_factor(ant_species),
         ant_species = fct_relevel(ant_species,
                                    "AB",
                                    "RRB",
                                    "BBR")) |> 
  mutate(species_name = case_match(
    ant_species,
    "AB" ~ "Crematogaster sjostedti",
    "RRB" ~ "Crematogaster mimosae",
    "BBR" ~ "Crematogaster nigriceps"
  )) 

slice_sample(nests_clean, n = 10)
```


## b. Response variable

For the response variable, case_control, the 1's signify there is a nest, of any bird species, present, while a 0 signifies there is no nest present. 

## c. Purpose of study

Tree height: Tree height is a continuous variable. As tree height increases, the probability of a nest occurrence would increase because of increased distance between nests and browsing mammals.

Acacia ant species: Acacia ant species is a categorical variable. 

## d. Table of models
| Model number | Tree Height | Ant Species | Interaction | 
|--------------|:-----------:|:-----------:|:-----------:| 
| 1            |      X      |      X      |     X       | 
| 2            |      X      |      X      |             |

## e. Run the models

```{r}
#| message: false
#| label: running models

# model 1: Model with interaction term

model1 <- glm(case_control ~ height_cm * ant_species,
              data = nests_clean,
              family = "binomial"
              )

# model 2: Model without interaction term

model2 <- glm(case_control ~ height_cm + ant_species,
              data = nests_clean,
              family = "binomial"
              )

```


## f. Select the best model

```{r}
#| label: model-selection

AICc(
model1,
model2
) |> 
arrange(AICc)
```

The best model as determined by Akaike's Information Criterion (AIC) was model 1, with interaction.

## g. Check the diagnostics

```{r}
#| label: check diagnostics
#| fig-height: 6
#| fig-width: 10

plot(
  simulateResiduals(model1)
)

```

## h. Visualize the model predictions

```{r}
#| label: Visualizing model predictions 1

# creating new object mod_preds
mod_preds <- ggpredict(model1,
                       #adding predictor variables
                       terms = c("height_cm",
                                 "ant_species")) |>
  #renaming group column ant_species
  dplyr::rename(ant_species = group)


#ggplot base
# x = height, y = nest occurrence, categorized by ant species
ggplot(nests_clean,
       aes( x = height_cm,
            y = case_control)) +
  # adding points representing observations
  geom_point(aes(color = ant_species),
             shape = 21) +
  # adding confidence level ribbon based on mod_preds
  geom_ribbon(data = mod_preds,
              aes(x = x,
                  y = predicted,
                  ymin = conf.low,
                  ymax = conf.high,
                  fill = ant_species,
              alpha = 0.1)) +
  # adding prediction line
  geom_line(data = mod_preds,
            aes( x = x,
                 y = predicted,
                 color = ant_species)) +
  # creating different panels
  # labelling by full name instead of abbreviations
  facet_wrap(~ant_species,
             labeller = labeller(
               ant_species = c(AB = "Crematogaster sjostedti",
                               RRB = "Crematogaster mimosae",
                               BBR = "Crematogaster nigriceps"))) +
  #custom colors
  scale_color_manual(values = c(
    "AB" = "skyblue",
    "RRB" = "turquoise4",
    "BBR" = "khaki3"
  )) +
  # matching ribbon colors
  scale_fill_manual(values = c(
  "AB" = "skyblue",
  "RRB" = "turquoise4",
  "BBR" = "khaki3"
)) +
  # adding x range
  # adding line breaks
  scale_x_continuous(limits = c(50, 600),
                     breaks = c(200, 400, 600)) +
  # getting rid of legend
  theme(legend.position = "none",
        panel.background = element_blank(), # removing grey background
        panel.grid.major = element_blank() # removing grid lines
        ) +
 # labelling axes
   labs(x = "Tree height (cm)",
       y = "Probability of nest occurrence")

```

## i. Write a caption for your figure

**Figure 1: As height increases, the probability of bird nest occurrence increases with each acacia ant species.** Data from nest_data_final.csv (Lujan E.et al, Zenodo, 2023). Lines represent predictions of probability of nest occurrence (lbs) given acacia species; Crematogaster sjostedti (AB), Crematogaster mimosae (RRB), Crematogaster nigriceps (BBR), (total n = 270) and tree height in cm. Meanwhile, points represent actual observations of if a nest occurred or not(0 = nest not present, 1 = nest present). The probability of nest occurrence increases as tree height increases. Nevertheless, the probability increases at different rates with each ant species, where Crematogaster sjostedti has the most gradual increase in probability followed by Crematogaster mimosae and Crematogaster nigriceps. Ribbons represent a 95% confidence interval. Colors represent acacia ant species (skyblue: Crematogaster sjostedti, teal: Crematogaster mimosae, khaki: Crematogaster nigriceps).

## j. Calculate model predictions

```{r}
#| label: calculate model predictions

ggpredict(model1,
          terms = c("height_cm [600]",
                    "ant_species"))

```

## k. Interpret your results

Acacia ant species and tree height (cm) interact to predict the probability of nest occurrence. As tree height increases, probability of nest occurrence increases.

Furthermore, at a tree height of 600 cm, the probability of nest occurrence is 0.20 for Crematogaster sjostedti (95% CI: [0.00, 0.99]), 0.74 for Crematogaster mimosae (95% CI: [0.48, 0.90]), and 1.00 for Crematogaster nigriceps (95% CI: [1.00, 1.00]). Indicating higher probability of nest occurrence for Crematogaster nigriceps at the highest tree height measured, 600 cm. 

# Problem 3. Affective and exploratory visualizations

## a. Comparing visualizations

There are many differences but a large difference is the amount of observations and data points, also of course they are displayed very differently. The affective visualization is a lot more artistic and its hard to tell it is data at first. Both graphs converge however in showing the low amount of caffeine consumed the more sleep in hours was had. Nevertheless, the patterns differ somewhat between the two. The graph from homework 02 did not have any distinct patterns and looked relatively random although there were a couple points that showed a pattern of negative correlation starting to emerge. The affective visualization had a much more prominent decrease and much more data to go off of. From week 9, I received mostly suggestions about coloring and incorporated these through watercolor for my final product. 

## b. Designing an analysis

Pearsons - measure strength of correlation between hours of sleep and ounces of caffeine consumed

## c. Check any assumptions and run your analysis

## d. Create a visualization

## e. Write a caption

## f. Write about your results

## g. Sharing your affective visualization

















